Exact Phylodynamics via Structured Markov Genealogy Processes

### 2024-01-20 ms.Rnw
edits

\begin{figure}
  \begin{center}
    \resizebox{0.9\linewidth}{!}{
      \begin{tikzpicture}[scale=1]
        \tikzstyle{box}=[color=black, draw, fill=white, very thick, minimum size=3em]
        \tikzstyle{label}=[font=\Large]
        \tikzstyle{coordinate}=[inner sep=0pt,outer sep=0pt]
        \tikzstyle{flow}=[color=black, very thick, >=stealth]
        \tikzstyle{deme}=[fill=black!20!white]
        \tikzstyle{modulate}=[color=blue, thick, >=latex]
        \coordinate (origin) at (0,0);
        \node[label] (lab) at ($(origin)+(-0.5,2.1)$) {\textbf{A}};
        \input{figs/seir_diagram}
        \coordinate (origin) at (9,0);
        \node[label] (lab) at ($(origin)+(-0.5,2.1)$) {\textbf{B}};
        \input{figs/siir_diagram}
        \coordinate (origin) at (0,-4);
        \node[label] (lab) at ($(origin)+(-0.5,2.1)$) {\textbf{C}};
        \input{figs/covid_diagram}
      \end{tikzpicture}
    }
  \end{center}
  \caption{
    Examples of discretely-structured population models.
    Demes are shaded.
    (A) An SEIRS model.
    Susceptible individuals ($\lab{S}$), once infected, enter a transient incubation phase ($\lab{E}$) before they become infectious ($\lab{I}$).
    Upon recovery ($\lab{R}$), individuals experience immunity from reinfection.
    If this immunity wanes, they re-enter the susceptible compartment.
    Pathogen lineages are to be found in hosts within the $\lab{E}$ and $\lab{I}$ compartments only.
    Accordingly, there are two demes ($\Demes=\Set{\lab{E},\lab{I}}$).
    If there is exactly one lineage per host, then the occupancy, $n(\Xr_t)=(n_{\lab{E}}(\Xr_t),n_{\lab{I}}(\Xr_t))$, is the integer 2-vector giving the numbers of hosts in the respective compartments.
    \AAK{[Perhaps another one or two examples here?]}
    \AAK{[We could add dots to the deme compartments to signify individuals\dots.]}
    \label{fig:examples}
  }
\end{figure}

We will assume that our population process is a time-inhomogeneous Markov jump process, $\Xr_t$, $t\in\Rp$, taking values in some space $\Xspace$.
In earlier work \citep{King2022}, we limited ourselves to the case $\Xspace=\Z^d$, but here we assume only that $\Xspace$ is a complete metric space with a countable dense subset, \ie a Polish space.
The population process is completely specified by its initial-state distribution, $p_0$, and its transition rates $\alpha$.
In particular, we suppose that
\begin{equation}\label{eq:ic}
  \Prob{\Xr_0\in\mathcal{E}}=\int_{\mathcal{E}}{p_0(x)\,\dd{x}}
\end{equation}
for all measurable sets $\mathcal{E}\subseteq\Xspace$.
For any $t\in\Rp$, $x,x'\in\Xspace$, we think of the quantity $\alpha(t,x,x')$ as the instantaneous hazard of a jump from $x$ to $x'$.
More precisely, the transition rates have the following properties:
\begin{equation*}
  \begin{gathered}
    \alpha(t,x,x')\ge{0}, \qquad \int_{\Xspace}{\alpha(t,x,x')\,\dd{x'}}<\infty,\\
  \end{gathered}
\end{equation*}
for all $t\in\Rp$ and $x,x'\in\Xspace$.
Henceforth, we understand that integrals are taken over all of $\Xspace$ unless otherwise specified.
Let $\Kr_t$ be the number of jumps that $\Xr$ has taken by time $t$.
We assume that $\Kr_t$ is a simple counting process so that
\begin{equation*}
  \begin{gathered}
    \CondProb{\Kr_{t+\Delta}=n+1}{\Kr_{t}=n}=\Delta\,\int{\alpha(t,x,x')\,\dd{x'}}+o(\Delta),\\
    \CondProb{\Kr_{t+\Delta}>n+1}{\Kr_{t}=n}=o(\Delta),\\
    \CondProb{\Xr_{t+\Delta}\in\mathcal{E}}{\Xr_{t}=x, \Kr_{t+\Delta}-\Kr_{t}=1}=\frac{\int_{\mathcal{E}}{\alpha(t,x,x')\,\dd{x'}}}{\int{\alpha(t,x,x')\,\dd{x'}}}.
  \end{gathered}
\end{equation*}
We will further assume that $\Xr_t$ is non-explosive, \ie that $\Prob{\Kr_t<\infty}=1$ for all $t$.

@@

The above may be compactly summarized by stating that if $w(t,x)$ satisfies the Kolmogorov forward equation (KFE),
\begin{equation}\label{eq:kfe}
  \frac{\partial{w}}{\partial{t}}(t,x)=\int\!w(t,x')\,\alpha(t,x',x)\,\dd{x'}-\int\!w(t,x)\,\alpha(t,x,x')\,\dd{x'},
\end{equation}
and if, moreover, $w(0,x) = p_0(x)$,
then $\int_{\mathcal{E}}\!{w(t,x)\,\dd{x}}=\Prob{\Xr_t\in\mathcal{E}}$ for every measurable $\mathcal{E}\subseteq{\Xspace}$.
\Cref{eq:kfe} is sometimes called the \emph{master equation} for $\Xr_t$.

@@

\begin{figure}
  \begin{center}
    \resizebox{0.9\linewidth}{!}{
      \begin{tikzpicture}[scale=1]
        \usetikzlibrary{shapes,arrows,positioning}
        \tikzstyle{coordinate}=[inner sep=0pt,outer sep=0pt]
        \tikzstyle{state}=[shape=ellipse, color=black, draw, font=\LARGE, fill=white, thick, minimum size=5em]
        \tikzstyle{trans}=[color=black, thick, >=stealth]
        \node[state] (base) at (0,0) {$x=(S,E,I,R)$};
        \node[state] (trans) at (165:8.4) {$x'=(S-1,E+1,I,R)$};
        \node[state] (prog) at (195:8) {$x'=(S,E-1,I+1,R+1)$};
        \node[state] (recov) at (295:4) {$x'=(S,E,I-1,R+1)$};
        \node[state] (wane) at (355:9) {$x'=(S+1,E,I,R-1)$};
        \node[state] (sample) at (18:7.5) {$x'=(S,E,I,R)$};
        \draw[trans,->] (base) -- (trans) node[midway,above,sloped] {$\lab{Trans}$};
        \draw[trans,->] (base) -- (prog) node[midway,above,sloped] {$\lab{Prog}$};
        \draw[trans,->] (base) -- (recov) node[midway,above,sloped] {$\lab{Recov}$};
        \draw[trans,->] (base) -- (wane) node[midway,above,sloped] {$\lab{Wane}$};
        \draw[trans,->] (base) -- (sample) node[midway,above,sloped] {$\lab{Sample}$};
        \node[font=\Large] (U) at (95:3.2) {$\Jumps=\Set{\lab{Trans},\lab{Prog},\lab{Recov},\lab{Wane},\lab{Sample}}$};
      \end{tikzpicture}
    }
  \end{center}
  \caption{
    Markov state transition diagram for the SEIRS model depicted in \cref{fig:examples}A.
    The state, $x$, is characterized by four numbers, $S$, $E$, $I$, and $R$.
    From a given state $x$, there are five possible kinds of jumps $x\mapsto{x'}$.
    Accordingly, the set, $\Jumps$, of jump marks has five elements.
    Each of these is of a different type:
    $\lab{Trans}$ (transmission) is of birth type,
    $\lab{Prog}$ (progression) is of migration type,
    $\lab{Recov}$ (recovery) is of death type,
    $\lab{Sample}$ (sampling) is of sample type,
    and $\lab{Wane}$ (loss or waning of immunity) is of neutral type.
    See \cref{sec:jump_types} for a description of these jump types.
    Note that, in this formulation, when a sampling event occurs, the state does not change.
    \label{fig:markov_state}
  }
\end{figure}

\subsubsection{Structured populations, demes}

@@

In the following, it will be useful to break the jumps into distinct categories.
For this purpose, we let $\Jumps$ be a countable set of jump \emph{marks} such that
\begin{equation*}
  \alpha(t,x,x')=\sum_u{\alpha_u(t,x,x')}.
\end{equation*}
In \cref{fig:markov_state}, we use the marks to distinguish events of different types.
Here and in the following, sums over $u$ are taken over the whole of $\Jumps$ unless otherwise indicated.

@@

Consider the Markov process $(\Xr_t,\Ur_t)$.
We define its \emph{history process}, $\Hr_t$, to be the restriction of the random function $s\mapsto(\Xr_s,\Ur_s)$ to the interval $[0,t]$.
Note that $\Hr_t$ is itself trivially a Markov process, since it contains its own history.
Alternatively, one can think of $\Hr_t$ as consisting of the sequence
$\left(\left(\Trh_k,\Xrh_k,\Urh_k\right)\right)_{k=0}^{\Kr_t}$.
In particular, conditional on $\Hr_t$, both $\Xr_t$ and $\Ur_t$ are deterministic as are $\Kr_t$ and the embedded chains, $\Xrh_k$, $\Urh_k$, and the point process of event times $\Trh_k$.
The probability measure, $\pi^{\mathrm{H}}$, for $\Hr_t$ can be expressed in terms of these:
\begin{equation*}\label{eq:Hdens}
  \begin{aligned}
    \pi^{\mathrm{H}}(\dd{\H_t})
    =&p_{0}(\Xh_{0})\,\dd{\Xh_0}\,
    \prod_{k=1}^{K_{t}}{\alpha_{\Uh_k}\!\!\left(\Th_k,\Xh_{k-1},\Xh_{k}\right)\,\dd{\Xh_k}\,\dd{\Th_k}}\\
    &\qquad\times\exp{\left(-\sum_{k=0}^{K_t}{\int_{\Th_{k}}^{\Th_{k+1}}{\sum_{u}{\int{\alpha_{u}(t',\Xh_{k},x')\,\dd{x'}}}}\,\dd{t'}}\right)},
  \end{aligned}
\end{equation*}
where again, by convention, $\Th_0=0$ and $\Th_{K_t+1}=t$.

\subsection{Binomial ratio}\label{sec:binomial_ratio}

@@

\subsection{Jump types}\label{sec:jump_types}

>>>>

\begin{figure}
  \begin{center}
    \resizebox{0.9\linewidth}{!}{
      \begin{tikzpicture}[scale=1]
        \colorlet{darkblue}{blue!50!black}
        \tikzstyle{box}=[draw=black, text=black, fill=white, very thick, minimum size=3em]
        \tikzstyle{ibox}=[draw=darkblue]
        \tikzstyle{deme}=[fill=black!20!white]
        \tikzstyle{label}=[font=\Large]
        \tikzstyle{coordinate}=[inner sep=0pt,outer sep=0pt]
        \tikzstyle{flow}=[draw=black, very thick, >=stealth]
        \tikzstyle{modulate}=[draw=darkblue, >=Circle]

        %% SEIRS model
        \coordinate (origin) at (0,0);
        \node[label] (lab) at (origin) {\textbf{A}};
        \node [box] (S) at ($(origin)+(1,-1)$) {$\lab{S}$};
        \node [box,deme] (E) at ($(S)+(2,0)$) {$\lab{E}$};
        \node [box,ibox,deme] (I) at ($(E)+(2,0)$) {$\lab{I}$};
        \node [box] (R) at ($(I)+(2,0)$) {$\lab{R}$};
        \coordinate (overR) at ($(R)+(0,1)$);
        \coordinate (midSE) at ($(S)!0.5!(E)$);
        \draw [flow,->] (S) -- (E);
        \draw [flow,->] (E) -- (I);
        \draw [flow,->] (I) -- (R);
        \draw [flow,->] (R) -- (overR) -- (S |- overR) -- (S);
        \draw [modulate,->] (I.north west) .. controls ($(I)+(-1,1)$) and ($(midSE)+(0,1)$) .. (midSE);
        \node[font=\normalsize] (demes) at ($(E)!0.5!(I)+(0,-1)$) {$\Demes=\{\lab{E},\lab{I}\}$};

        %% SIIR (two-strain) model
        \coordinate (origin) at (0,-3);
        \node[label] (lab) at (origin) {\textbf{B}};
        \node [box] (S) at ($(origin)+(1,-2)$) {$\lab{S}$};
        \node [box] (R1) at ($(S)+(6,1)$) {$\lab{R}_1$};
        \node [box] (R2) at ($(S)+(6,-1)$) {$\lab{R}_2$};
        \coordinate (overR) at ($(R1)+(0,1)$);
        \coordinate (underR) at ($(R2)-(0,1)$);
        \node [box,deme] (E1) at ($(S)+(2,1)$) {$\lab{E}_1$};
        \node [box,deme] (E2) at ($(S)+(2,-1)$) {$\lab{E}_2$};
        \node [box,ibox,deme] (I1) at ($(E1)+(2,0)$) {$\lab{I}_1$};
        \node [box,ibox,deme] (I2) at ($(E2)+(2,0)$) {$\lab{I}_2$};
        \draw [flow,->] (S) -- (E1);
        \draw [flow,->] (E1) -- (I1);
        \draw [flow,->] (I1) -- (R1);
        \draw [flow,->] (S) -- (E2);
        \draw [flow,->] (E2) -- (I2);
        \draw [flow,->] (I2) -- (R2);
        \draw [flow,->] (R1) -- (overR) -- (S |- overR) -- (S);
        \draw [flow,->] (R2) -- (underR) -- (S |- underR) -- (S);
        \coordinate (midSE1) at ($(S)!0.5!(E1)$);
        \coordinate (midSE2) at ($(S)!0.5!(E2)$);
        \draw [modulate,->] (I1.north west) .. controls ($(I1)+(-1,1)$) and ($(midSE1)+(0,1.8)$) .. (midSE1);
        \draw [modulate,->] (I2.south west) .. controls ($(I2)+(-1,-1)$) and ($(midSE2)+(0,-1.8)$) .. (midSE2);
        \node[font=\normalsize] (demes) at ($(E2)!0.5!(I2)+(0,-2)$) {$\Demes=\{\lab{E}_1,\lab{I}_1,\lab{E}_2,\lab{I}_2\}$};

        %% COVID model
        \coordinate (origin) at (9,0);
        \node[label] (lab) at (origin) {\textbf{C}};
        \node [box] (S) at ($(origin)+(1,-2)$) {$\lab{S}$};
        \node [box] (R) at ($(S)+(7,1)$) {$\lab{R}$};
        \coordinate (overR) at ($(R)+(0,1)$);
        \node [box,deme] (E) at ($(S)+(2,0)$) {$\lab{E}$};
        \node [box,ibox,deme] (Ia) at ($(E)+(2,1)$) {$\lab{I_A}$};
        \node [box,ibox,deme] (Is) at ($(E)+(2,-1)$) {$\lab{I_S}$};
        \node [box] (H) at ($(Is)+(2,0)$) {$\lab{H}$};
        \node [box] (D) at ($(H)+(2,0)$) {$\lab{D}$};
        \draw [flow,->] (S) -- (E);
        \draw [flow,->] (E) -- (Ia.west);
        \draw [flow,->] (E) -- (Is.west);
        \draw [flow,->] (Ia) -- (R);
        \draw [flow,->] (Is) -- (H);
        \draw [flow,->] (H) -- (R);
        \draw [flow,->] (Is) -- (R);
        \draw [flow,->] (H) -- (D);
        \draw [flow,->] (R) -- (overR) -- (S |- overR) -- (S);
        \coordinate (midSE) at ($(S)!0.5!(E)$);
        \draw [modulate,->] (Ia.north west) .. controls ($(Ia.north west)+(165:1)$) and ($(midSE)+(0,2)$) .. (midSE);
        \draw [modulate,->] (Is.south west) .. controls ($(Is.south west)+(195:1)$) and ($(midSE)+(0,-2)$) .. (midSE);
        \node[font=\normalsize] (demes) at ($(Is)+(0,-1)$) {$\Demes=\{\lab{E},\lab{I_A},\lab{I_S}\}$};

        %% SI2R (super-spreader) model
        \coordinate (origin) at (9,-5);
        \node[label] (lab) at (origin) {\textbf{D}};
        \node [box] (S) at ($(origin)+(2,-1)$) {$\lab{S}$};
        \node [box,deme] (E) at ($(S)+(2,0)$) {$\lab{E}$};
        \node [box,ibox,deme] (Il) at ($(E)+(2,0)$) {$\lab{I_L}$};
        \node [box,ibox,deme] (Ih) at ($(Il)+(0,-2)$) {$\lab{I_H}$};
        \node [box] (R) at ($(Il)+(2,0)$) {$\lab{R}$};
        \coordinate (overR) at ($(R)+(0,1)$);
        \coordinate (midSE) at ($(S)!0.5!(E)$);
        \draw [flow,->] (S) -- (E);
        \draw [flow,->] (E) -- (Il);
        \draw [flow,<->] (Il) -- (Ih);
        \draw [flow,->] (Il) -- (R);
        \draw [flow,->] (Ih.east) -- (R |- Ih.east) -- (R);
        \draw [flow,->] (R) -- (overR) -- (S |- overR) -- (S);
        \draw [modulate,->] (Il.north west) .. controls ($(Il.north west)+(135:0.8)$) and ($(midSE)+(0,1)$) .. (midSE);
        \draw [modulate,->] (Ih.west) .. controls ($(Ih.west)+(180:0.8)$) and ($(midSE)+(0,-1)$) .. (midSE);
        \node[font=\normalsize] (demes) at (midSE |- Ih) {$\Demes=\{\lab{E},\lab{I_L},\lab{I_H}\}$};
      \end{tikzpicture}
    }
  \end{center}
  \caption{
    Examples of discretely-structured population models.
    Demes are shaded.
    \textbf{(A)} An SEIRS model.
    Susceptible individuals ($\lab{S}$), once infected, enter a transient incubation phase ($\lab{E}$) before they become infectious ($\lab{I}$).
    Upon recovery ($\lab{R}$), individuals experience immunity from reinfection.
    If this immunity wanes, they re-enter the susceptible compartment.
    Pathogen lineages are to be found in hosts within the $\lab{E}$ and $\lab{I}$ compartments only.
    Accordingly, there are two demes ($\Demes=\Set{\lab{E},\lab{I}}$).
    If there is exactly one lineage per host, then the occupancy, $n(\Xr_t)=(n_{\lab{E}}(\Xr_t),n_{\lab{I}}(\Xr_t))$, is the integer 2-vector giving the numbers of hosts in the respective compartments.
    \AAK{[We could add dots to the deme compartments to signify individuals\dots.]}
    \label{fig:examples}
  }
\end{figure}

We will assume that our population process is a time-inhomogeneous Markov jump process, $\Xr_t$, $t\in\Rp$, taking values in some space $\Xspace$.
In earlier work \citep{King2022}, we limited ourselves to the case $\Xspace=\Z^d$, but here we assume only that $\Xspace$ is a complete metric space with a countable dense subset, \ie a Polish space.
The population process is completely specified by its initial-state distribution, $p_0$, and its transition rates $\alpha$.
In particular, we suppose that
\begin{equation}
  \label{eq:ic}
  \Prob{\Xr_0\in\mathcal{E}}=\int_{\mathcal{E}}{p_0(x)\,\dd{x}}
\end{equation}
for all measurable sets $\mathcal{E}\subseteq\Xspace$.
For any $t\in\Rp$, $x,x'\in\Xspace$, we think of the quantity $\alpha(t,x,x')$ as the instantaneous hazard of a jump from $x$ to $x'$.
More precisely, the transition rates have the following properties:
\begin{equation*}
  \begin{gathered}
    \alpha(t,x,x')\ge{0}, \qquad \int_{\Xspace}{\alpha(t,x,x')\,\dd{x'}}<\infty,\\
  \end{gathered}
\end{equation*}
for all $t\in\Rp$ and $x,x'\in\Xspace$.
Henceforth, we understand that integrals are taken over all of $\Xspace$ unless otherwise specified.
Let $\Kr_t$ be the number of jumps that $\Xr$ has taken by time $t$.
We assume that $\Kr_t$ is a simple counting process so that
\begin{equation*}
  \begin{gathered}
    \CondProb{\Kr_{t+\Delta}=n+1}{\Kr_{t}=n}=\Delta\,\int{\alpha(t,x,x')\,\dd{x'}}+o(\Delta),\\
    \CondProb{\Kr_{t+\Delta}>n+1}{\Kr_{t}=n}=o(\Delta),\\
    \CondProb{\Xr_{t+\Delta}\in\mathcal{E}}{\Xr_{t}=x, \Kr_{t+\Delta}-\Kr_{t}=1}=\frac{\int_{\mathcal{E}}{\alpha(t,x,x')\,\dd{x'}}}{\int{\alpha(t,x,x')\,\dd{x'}}}.
  \end{gathered}
\end{equation*}
We will further assume that $\Xr_t$ is non-explosive, \ie that $\Prob{\Kr_t<\infty}=1$ for all $t$.

@@

The above may be compactly summarized by stating that if $w(t,x)$ satisfies the Kolmogorov forward equation (KFE),
\begin{equation}
  \label{eq:kfe}
  \frac{\partial{w}}{\partial{t}}(t,x)=\int\!w(t,x')\,\alpha(t,x',x)\,\dd{x'}-\int\!w(t,x)\,\alpha(t,x,x')\,\dd{x'},
\end{equation}
and if, moreover, $w(0,x) = p_0(x)$,
then $\int_{\mathcal{E}}\!{w(t,x)\,\dd{x}}=\Prob{\Xr_t\in\mathcal{E}}$ for every measurable $\mathcal{E}\subseteq{\Xspace}$.
\Cref{eq:kfe} is sometimes called the \emph{master equation} for $\Xr_t$.

@@

\begin{figure}
  \begin{center}
    \resizebox{0.9\linewidth}{!}{
      \begin{tikzpicture}[scale=1]
        \usetikzlibrary{shapes,arrows,positioning}
        \tikzstyle{coordinate}=[inner sep=0pt,outer sep=0pt]
        \tikzstyle{state}=[shape=ellipse, color=black, draw, font=\LARGE, fill=white, thick, minimum height=8em]
        \tikzstyle{trans}=[color=black, font=\Large, thick, >=stealth]
        \node[state] (base) at (0,0) {$x=(S,E,I,R)$};
        \node[state] (trans) at (165:8) {$x'=(S-1,E+1,I,R)$};
        \node[state] (prog) at (205:7) {$x'=(S,E-1,I+1,R+1)$};
        \node[state] (recov) at (305:5) {$x'=(S,E,I-1,R+1)$};
        \node[state] (wane) at (350:8) {$x'=(S+1,E,I,R-1)$};
        \node[state] (sample) at (18:7.5) {$x'=(S,E,I,R)$};
        \draw[trans,->] (base) -- (trans) node[midway,above,sloped] {$\lab{Trans}$};
        \draw[trans,->] (base) -- (prog) node[midway,above,sloped] {$\lab{Prog}$};
        \draw[trans,->] (base) -- (recov) node[midway,above,sloped] {$\lab{Recov}$};
        \draw[trans,->] (base) -- (wane) node[midway,above,sloped] {$\lab{Wane}$};
        \draw[trans,->] (base) -- (sample) node[midway,above,sloped] {$\lab{Sample}$};
        \node[font=\LARGE] (U) at (95:3.5) {$\Jumps=\Set{\lab{Trans},\lab{Prog},\lab{Recov},\lab{Wane},\lab{Sample}}$};
      \end{tikzpicture}
    }
  \end{center}
  \caption{
    Markov state transition diagram for the SEIRS model depicted in \cref{fig:examples}A.
    The state, $x$, is characterized by four numbers, $S$, $E$, $I$, and $R$.
    From a given state $x$, there are five possible kinds of jumps $x\mapsto{x'}$.
    Accordingly, the set, $\Jumps$, of jump marks has five elements.
    Each of these is of a different type:
    $\lab{Trans}$ (transmission) is of birth type,
    $\lab{Prog}$ (progression) is of migration type,
    $\lab{Recov}$ (recovery) is of death type,
    $\lab{Sample}$ (sampling) is of sample type,
    and $\lab{Wane}$ (loss or waning of immunity) is of neutral type.
    See \cref{sec:jump_types} for a description of these jump types.
    Note that, in this formulation, when a sampling event occurs, the state does not change.
    \label{fig:markov_state}
  }
\end{figure}

\subsubsection{Structured populations, demes, and deme occupancy}
\label{sec:demes}

@@

In the following, it will be useful to break the jumps into distinct categories.
For this purpose, we let $\Jumps$ be a countable set of jump \emph{marks} such that
\begin{equation*}
  \alpha(t,x,x')=\sum_u{\alpha_u(t,x,x')}.
\end{equation*}
In \cref{fig:markov_state}, we use the marks to distinguish events with distinct effects on genealogical relationships.
Here and in the following, sums over $u$ are taken over the whole of $\Jumps$ unless otherwise indicated.

@@

Consider the Markov process $(\Xr_t,\Ur_t)$.
We define its \emph{history process}, $\Hr_t$, to be the restriction of the random function $s\mapsto(\Xr_s,\Ur_s)$ to the interval $[0,t]$.
Note that $\Hr_t$ is itself trivially a Markov process, since it contains its own history.
Alternatively, one can think of $\Hr_t$ as consisting of the sequence
$\left(\left(\Trh_k,\Xrh_k,\Urh_k\right)\right)_{k=0}^{\Kr_t}$.
In particular, conditional on $\Hr_t$, both $\Xr_t$ and $\Ur_t$ are deterministic as are $\Kr_t$ and the embedded chains, $\Xrh_k$, $\Urh_k$, and the point process of event times $\Trh_k$.
The probability measure, $\pi^{\mathrm{H}}$, for $\Hr_t$ can be expressed in terms of these:
\begin{equation*}
  \label{eq:Hdens}
  \begin{aligned}
    \pi^{\mathrm{H}}(\dd{\H_t})
    =&p_{0}(\Xh_{0})\,\dd{\Xh_0}\,
    \prod_{k=1}^{K_{t}}{\alpha_{\Uh_k}\!\!\left(\Th_k,\Xh_{k-1},\Xh_{k}\right)\,\dd{\Xh_k}\,\dd{\Th_k}}\\
    &\qquad\times\exp{\left(-\sum_{k=0}^{K_t}{\int_{\Th_{k}}^{\Th_{k+1}}{\sum_{u}{\int{\alpha_{u}(t',\Xh_{k},x')\,\dd{x'}}}}\,\dd{t'}}\right)},
  \end{aligned}
\end{equation*}
where again, by convention, $\Th_0=0$ and $\Th_{K_t+1}=t$.

\subsection{Binomial ratio}
\label{sec:binomial_ratio}

@@

\subsection{Jump types}
\label{sec:jump_types}

\begin{figure}
  \begin{center}
    \resizebox{0.8\linewidth}{!}{
      \begin{tikzpicture}[scale=2]
        \usetikzlibrary{shapes,arrows,positioning}
        \usetikzlibrary{arrows.meta,positioning,decorations}
        \colorlet{darkblue}{blue!50!black}
        \colorlet{maize}{yellow!90!black}
        \colorlet{grey}{black!30!white}
        \colorlet{darkgreen}{green!40!black}
        \colorlet{lightblue}{white!30!blue}
        \tikzstyle{label}=[font=\Large]
        \tikzstyle{coordinate}=[inner sep=0pt,outer sep=0pt]
        \tikzstyle{deme1}=[color=grey, line width=3pt, >={Circle[width=5pt,length=5pt]}]
        \tikzstyle{deme2}=[color=darkblue, line width=3pt, >={Circle[width=5pt,length=5pt]}]
        \tikzstyle{deme3}=[color=maize, line width=3pt, >={Circle[width=5pt,length=5pt]}]
        \tikzstyle{axis}=[color=black, thick, >=stealth]
        \tikzstyle{timepoint}=[color=red!70!black]

        \def\ysp{0.2} %% spacing between adjacent lineages

        %% Birth-type event
        \begin{scope}[shift={(0,0)}]
          \coordinate (origin) at (0,0);
          \node[label] (lab) at (origin) {\textbf{A}};
          \coordinate (left) at ($(origin)+(0.1,-0.1)$);
          \coordinate (right) at ($(left)+(1,0)$);
          \coordinate (mid) at ($(left)!0.5!(right)$);
          \coordinate (rlab) at ($(right)+(0.1,0)$);
          \coordinate (bleft) at ($(left)+(0,-8*\ysp)$);
          \coordinate (bright) at (rlab |- bleft);
          \coordinate (temp) at ($(origin)+(0,-4*\ysp)$);
          \coordinate (temp1) at ($(temp)+(0,\ysp)$);
          \coordinate (temp2) at ($(temp)-(0,\ysp)$);
          \draw[deme2] (mid |- temp) -- (mid |- temp1) -- (right |- temp1);
          \draw[deme3] (mid |- temp) -- (mid |- temp2) -- (right |- temp2);
          \draw[deme1] (left |- temp) -- (right |- temp);
          \node[text=darkgreen,font=\Huge] at (mid |- temp) {$\bullet$};
          \foreach \y in {1,7} {
            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
            \draw[deme1] (left |- temp) -- (right |- temp);
          }
          \foreach \y in {6,8} {
            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
            \draw[deme2] (left |- temp) -- (right |- temp);
          }
          \foreach \y in {2,7} {
            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
            \draw[deme3] (left |- temp) -- (right |- temp);
          }
          \draw[timepoint] (mid) -- (mid |- bleft);
          \draw[axis,->] (bleft) -- (bright);
          \node[xshift=1ex] (tlab) at (bright.east) {$t$};
        \end{scope}

        %% Death-type event
        \begin{scope}[shift={(2,0)}]
          \coordinate (origin) at (0,0);
          \node[label] (lab) at (origin) {\textbf{B}};
          \coordinate (left) at ($(origin)+(0.1,-0.1)$);
          \coordinate (right) at ($(left)+(1,0)$);
          \coordinate (mid) at ($(left)!0.5!(right)$);
          \coordinate (rlab) at ($(right)+(0.1,0)$);
          \coordinate (bleft) at ($(left)+(0,-8*\ysp)$);
          \coordinate (bright) at (rlab |- bleft);
          \foreach \y in {1,2,6} {
            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
            \draw[deme1] (left |- temp) -- (right |- temp);
          }
          \foreach \y in {3,8} {
            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
            \draw[deme2] (left |- temp) -- (right |- temp);
          }
          \coordinate (temp) at ($(origin)+(0,-4*\ysp)$);
          \draw[deme3,dashed] (left |- temp) -- (mid |- temp);
          \node[text=red,font=\Huge] at (mid |- temp) {$\times$};
          \foreach \y in {5,7} {
            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
            \draw[deme3] (left |- temp) -- (right |- temp);
          }
          \draw[timepoint] (mid) -- (mid |- bleft);
          \draw[axis,->] (bleft) -- (bright);
          \node[xshift=1ex] (tlab) at (bright.east) {$t$};
        \end{scope}

        %% Migration-type event
        \begin{scope}[shift={(4,0)}]
          \coordinate (origin) at (0,0);
          \node[label] (lab) at (origin) {\textbf{C}};
          \coordinate (left) at ($(origin)+(0.1,-0.1)$);
          \coordinate (right) at ($(left)+(1,0)$);
          \coordinate (mid) at ($(left)!0.5!(right)$);
          \coordinate (rlab) at ($(right)+(0.1,0)$);
          \coordinate (bleft) at ($(left)+(0,-8*\ysp)$);
          \coordinate (bright) at (rlab |- bleft);
          \foreach \y in {2,3,8} {
            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
            \draw[deme1] (left |- temp) -- (right |- temp);
          }
          \coordinate (temp) at ($(origin)+(0,-4*\ysp)$);
          \draw[deme2] (left |- temp) -- (mid |- temp);
          \draw[deme3] (mid |- temp) -- (right |- temp);
          \node[text=darkgreen,font=\Huge] at (mid |- temp) {$\bullet$};
          \foreach \y in {1,5} {
            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
            \draw[deme2] (left |- temp) -- (right |- temp);
          }
          \foreach \y in {6,7} {
            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
            \draw[deme3] (left |- temp) -- (right |- temp);
          }
          \draw[timepoint] (mid) -- (mid |- bleft);
          \draw[axis,->] (bleft) -- (bright);
          \node[xshift=1ex] (tlab) at (bright.east) {$t$};
        \end{scope}

        %% Sample event
        \begin{scope}[shift={(6,0)}]
          \coordinate (origin) at (0,0);
          \node[label] (lab) at (origin) {\textbf{D}};
          \coordinate (left) at ($(origin)+(0.1,-0.1)$);
          \coordinate (right) at ($(left)+(1,0)$);
          \coordinate (mid) at ($(left)!0.5!(right)$);
          \coordinate (rlab) at ($(right)+(0.1,0)$);
          \coordinate (bleft) at ($(left)+(0,-8*\ysp)$);
          \coordinate (bright) at (rlab |- bleft);
          \foreach \y in {2,5} {
            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
            \draw[deme1] (left |- temp) -- (right |- temp);
          }
          \foreach \y in {1,4,7} {
            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
            \draw[deme2] (left |- temp) -- (right |- temp);
          }
          \foreach \y in {3,6,8} {
            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
            \draw[deme3] (left |- temp) -- (right |- temp);
          }
          \coordinate (temp) at ($(origin)+(0,-2*\ysp)$);
          \node[text=lightblue,font=\Huge] at (mid |- temp) {$\bullet$};
          \coordinate (temp) at ($(origin)+(0,-6*\ysp)$);
          \node[text=lightblue,font=\Huge] at (mid |- temp) {$\bullet$};
          \draw[timepoint] (mid) -- (mid |- bleft);
          \draw[axis,->] (bleft) -- (bright);
          \node[xshift=1ex] (tlab) at (bright.east) {$t$};
        \end{scope}

        %% Neutral-type event
        \begin{scope}[shift={(0,-2)}]
          \coordinate (origin) at (0,0);
          \node[label] (lab) at (origin) {\textbf{E}};
          \coordinate (left) at ($(origin)+(0.1,-0.1)$);
          \coordinate (right) at ($(left)+(1,0)$);
          \coordinate (mid) at ($(left)!0.5!(right)$);
          \coordinate (rlab) at ($(right)+(0.1,0)$);
          \coordinate (bleft) at ($(left)+(0,-8*\ysp)$);
          \coordinate (bright) at (rlab |- bleft);
          \foreach \y in {4,6,7} {
            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
            \draw[deme1] (left |- temp) -- (right |- temp);
          }
          \foreach \y in {1,2,5} {
            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
            \draw[deme2] (left |- temp) -- (right |- temp);
          }
          \foreach \y in {3,8} {
            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
            \draw[deme3] (left |- temp) -- (right |- temp);
          }
          \draw[timepoint] (mid) -- (mid |- bleft);
          \draw[axis,->] (bleft) -- (bright);
          \node[xshift=1ex] (tlab) at (bright.east) {$t$};
        \end{scope}
